Temporal ordering and registration of images in studies of developmental dynamics

### 2014-05-27 image_analysis_paper/si_information.tex
edited paper; added references, changed figure

@@

\begin{figure}[H]
\includegraphics[height=9cm]{raw_data3b}
\caption{Images of cellularizing {\it Drosophila} embryos. Each image is of a different embryo in a different rotational orientation. $48$ of the images are of wild type embryos, and $42$ of the images are of mutant embryos.}
\end{figure}

@@

\begin{figure}[H]
\raisebox{5.2cm}{A}
\includegraphics[width=16.8cm]{data2_PCA_ordered}

\raisebox{3.7cm}{B}
\includegraphics[width=8.4cm]{data2_PCA_variance}
\raisebox{3.7cm}{C}
\includegraphics[width=8.4cm]{data2_PCA_proj}

@@

In general, the embedding coordinates are given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$, 
%
If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate.
%
As the first diffusion maps coordinate is constant, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate. 
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%
The $j^{th}$ embedding coordinate will be given by $\psi_{k,l}$ such that $0 < k \le d, jd +1 < l \le (j+1)d$,
and we select the $(k,l)$ pair such that the coordinate $\psi_{k, l}$ has the largest variability. 
%

@@

\begin{figure}[H]
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_1}\\
\includegraphics[width=\textwidth]{sphere2_1}\\
A
\end{minipage}
%
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_2}\\
\includegraphics[width=\textwidth]{sphere2_2}\\
B
\end{minipage}
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_3}\\
\includegraphics[width=\textwidth]{sphere2_3}\\
C
\end{minipage}
\begin{minipage}{0.2\textwidth} 
\centering
\includegraphics[width=\textwidth]{sphere_4}\\
\includegraphics[width=\textwidth]{sphere2_4}\\
D
\end{minipage}
\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions. {\it (A)} The original image. {\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image. {\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation. {\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation. The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
\label{fig:SO3_picture}
\end{figure}

@@

\subsection{Algorithms}

\subsubsection{Angular synchronization}

\begin{enumerate}
\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
\item Construct the matrix $H$ as in \eqref{eq:H_to_R}, which is a block matrix with entries $R_{ij}$.
\item Compute the top $2$ (for rotations) or $3$ (for rotations + translations) eigenvectors of $H$.
\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}
\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
\end{enumerate}

\subsubsection{Diffusion Maps}

\begin{enumerate}
\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}
\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}
\item Compute the top few eigenvectors of $A$, ordered with decreasing eigenvalues
\item If we only want to order the data, then sort the data by the entries in the second eigenvector.
\item For higher-dimensional embeddings (e.g., the mutants), use the entries of successive eigenvectors (the second eigenvector, the third eigenvector, etc.) as the embedding coordinates
\end{enumerate}

\subsubsection{Vector Diffusion Maps}

\begin{enumerate}
\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
%
\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
%
\item Construct the matrix $H$ as in \eqref{eq:H_low_rank}.
%
\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}, using the minimum distance obtained for each pair in step 1 as the pairwise distances in the Gaussian kernel.
%
\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}.
%
\item Construct the matrix $S$ as in \eqref{eq:vdm_S}.
%
\item Compute the top few eigenvectors of $S$.
%
\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
%
\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}.
%
\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
%
\item To compute the first (non-trivial) embedding coordinate for the data, compute $\psi_{k,l}$ as in \eqref{eq:vdm_coord}, taking $k=1, 2$ and $l=3, 4$ for the rotation case, or $k = 1, 2, 3$ and $l = 4, 5, 6$ for the rotation + translation case.
%
Select the $(k, l)$ coordinate which has the largest variance as the embedding coordinate.
%
\item To compute subsequent coordinates, use lower blocks for $l$ (e.g., for the next coordinate, under translations and rotations, use $k=1, 2, 3$ and $l = 7, 8, 9$). Again, compute $\psi_{k, l}$ as in \eqref{eq:vdm_coord}, and select the $(k, l)$ coordinate with largest variance.

\end{enumerate}

@@

I computed the FBsPCA basis for the set of images (using only the dpERK signal).
%
I then computed the bispectrum coefficients for each image expanded in this basis, and used these coefficients in a DMAPS calculation.
%
The results are shown below.
%
However, because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.

\begin{figure}[H]
\includegraphics[width=7cm]{FBsPCA_images}
\includegraphics[width=10cm]{FBsPCA_dmaps}
\caption{(Left) Images used for FBsPCA. (Right) First 4 dmaps variables, computed from the bispectrum coefficeints, plotted against the membrane thickness. Note there is no appreciable correlation: each dmaps variable simply selects one image from the data set as ``abnormal''.}
\end{figure}

>>>>

\renewcommand\thefigure{S\arabic{figure}}

@@

\begin{figure}[H]
\includegraphics[height=9cm]{raw_data3b}
\caption{Images of cellularizing {\it Drosophila} embryos. Each image is of a different embryo in a different rotational orientation. Half of the images are of wild type embryos, and half of the images are of mutant embryos.}
\end{figure}

@@

\begin{figure}[H]
\raisebox{5.2cm}{\sffamily{A}}
\includegraphics[width=16.8cm]{data2_PCA_ordered}

\raisebox{3.7cm}{\sffamily{B}}
\includegraphics[width=8.4cm]{data2_PCA_variance}
\raisebox{3.7cm}{\sffamily{C}}
\includegraphics[width=8.4cm]{data2_PCA_proj}

@@

In general, the embedding coordinates are given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$, 
%
If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate.
%
As the first diffusion maps coordinate is constant, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate. 
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%
The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is
$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2 $$

@@

\begin{figure}[H]
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_1}\\
\includegraphics[width=\textwidth]{sphere2_1}\\
\sffamily{A}
\end{minipage}
%
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_2}\\
\includegraphics[width=\textwidth]{sphere2_2}\\
\sffamily{B}
\end{minipage}
\begin{minipage}{0.2\textwidth}
\centering
\includegraphics[width=\textwidth]{sphere_3}\\
\includegraphics[width=\textwidth]{sphere2_3}\\
\sffamily{C}
\end{minipage}
\begin{minipage}{0.2\textwidth} 
\centering
\includegraphics[width=\textwidth]{sphere_4}\\
\includegraphics[width=\textwidth]{sphere2_4}\\
\sffamily{D}
\end{minipage}
\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions. {\it (A)} The original image. {\it (B)} Rotation around the x-axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image. {\it (C)} Rotation around the y-axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation. {\it (D)} Rotation around the z-axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation. The top row shows the three-dimensional spheres, and the bottom row shows the (two-dimensional) surface of the sphere in which we are interested.}
\label{fig:SO3_picture}
\end{figure}

@@

%\subsection{Algorithms}
%
%\subsubsection{Angular synchronization}
%
%\begin{enumerate}
%\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
%\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
%\item Construct the matrix $H$ as in \eqref{eq:H_to_R}, which is a block matrix with entries $R_{ij}$.
%\item Compute the top $2$ (for rotations) or $3$ (for rotations + translations) eigenvectors of $H$.
%\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
%\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}
%\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
%\end{enumerate}
%
%\subsubsection{Diffusion Maps}
%
%\begin{enumerate}
%\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}
%\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}
%\item Compute the top few eigenvectors of $A$, ordered with decreasing eigenvalues
%\item If we only want to order the data, then sort the data by the entries in the second eigenvector.
%\item For higher-dimensional embeddings (e.g., the mutants), use the entries of successive eigenvectors (the second eigenvector, the third eigenvector, etc.) as the embedding coordinates
%\end{enumerate}
%
%\subsubsection{Vector Diffusion Maps}
%
%\begin{enumerate}
%\item Compute the pairwise rotations (described in \eqref{eq:opt_angle}) or rotations + translations (described in \eqref{eq:opt_pairwise}) needed to register every pair of data points.
%%
%\item Construct the corresponding rotation matrices $R_{ij}$ for every pair $i, j$ of images (see \eqref{eq:R_theta} for rotations or \eqref{eq:total_rot} for rotations + translations).
%%
%\item Construct the matrix $H$ as in \eqref{eq:H_low_rank}.
%%
%\item Construct the matrix $W$ as in \eqref{eq:dmaps_W}, using the minimum distance obtained for each pair in step 1 as the pairwise distances in the Gaussian kernel.
%%
%\item Construct the matrix $A$ as in \eqref{eq:dmaps_A}.
%%
%\item Construct the matrix $S$ as in \eqref{eq:vdm_S}.
%%
%\item Compute the top few eigenvectors of $S$.
%%
%\item Recover the estimates $\hat{R}_1, \dots, \hat{R}_m$ from the top 2 or 3 eigenvectors as in \eqref{eq:R_hat}.
%%
%\item Project these estimates onto the closest proper rotation matrix using \eqref{eq:R_est}.
%%
%\item Recover the optimal rotation angle $\theta_i$ using \eqref{eq:R_theta} or rotation + translation using \eqref{eq:angle_invert} for each image
%%
%\item To compute the first (non-trivial) embedding coordinate for the data, compute $\psi_{k,l}$ as in \eqref{eq:vdm_coord}, taking $k=1, 2$ and $l=3, 4$ for the rotation case, or $k = 1, 2, 3$ and $l = 4, 5, 6$ for the rotation + translation case.
%%
%Select the $(k, l)$ coordinate which has the largest variance as the embedding coordinate.
%%
%\item To compute more coordinates, use the subsequent blocks for $l$ (e.g., for the next coordinate, under translations and rotations, use $k=1, 2, 3$ and $l = 7, 8, 9$). Again, compute $\psi_{k, l}$ as in \eqref{eq:vdm_coord}, and select the $(k, l)$ coordinate with largest variance.
%
%\end{enumerate}

@@

I computed the FBsPCA basis for the set of images (separately for the red and green signals).
%
I then computed the bispectrum coefficients for each image expanded in the FBsPCA basis, and used these coefficients in a DMAPS calculation.
%
The results are shown below.
%
However, because the bispectrum is not stable to deformations, it does not seem to be able to identify neighboring images.

\begin{figure}[H]
\raisebox{7.6cm}{\sffamily{A}}
\includegraphics[width=8cm]{raw_data1}
\hfill
\raisebox{7.6cm}{\sffamily{B}}
\includegraphics[width=7cm]{FBsPCA_evals}

\raisebox{2.2cm}{\sffamily{C}}
\hspace{-1cm}
\includegraphics[width=4cm, angle=90]{FBsPCA_dmaps}
\caption{(A) Images used for FBsPCA. (B) Eigenvalue spectrum from dmaps, using bispectrum coefficients. (C) First 4 dmaps variables, computed from the bispectrum coefficeints, plotted against the membrane thickness. Note there is no appreciable correlation: each dmaps variable simply selects one image from the data set as ``abnormal''.}
\end{figure}
